---
title: "Data Analysis Assignment"
author: "Jeeva"
date: "2023-09-10"
output: html_document
---

  
  # Introduction
  
  This is my data analysis assignment, where I will analyze the "insurance.csv" dataset using R.

## Dataset

I have loaded the dataset into a variable named 'df' for data frame.

```{r}
# Load the dataset into a variable (e.g., df for data frame)
df <- read.csv("insurance.csv")

# Descriptive statistics for quantitative variables
summary(df$charges)
summary(df$age)
summary(df$bmi)

# Descriptive statistics for categorical variables
table(df$sex)
table(df$children)
table(df$smoker)
table(df$region)

# Logarithmic transformation for 'charges'
df$log_charges <- log(df$charges)

# Descriptive statistics for the transformed variable 'log_charges'
summary(df$log_charges)

```{r}
# Load the ggplot2 library if not already loaded
library(ggplot2)

# Histogram for the 'age' variable
ggplot(df, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  labs(title = "Distribution of Age", x = "Age", y = "Frequency")


# Scatterplot between 'charges' and 'age'
ggplot(df, aes(x = charges, y = age)) +
  geom_point()


